Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood Constraints

### 2021-01-19 paper/paper.tex
More things

\usepackage{showframe}

@@

\author{
Ciaran McCreesh$^1$
\and
Sonja Kraiczy$^2$
\affiliations
$^1$University of Glasgow\\
$^2$University of Oxford\\
\emails
ciaran.mccreesh@glasgow.ac.uk
}

@@

A \emph{graph homomorphism} is a function between two graphs that maps adjacent vertices to adjacent
vertices. An injective graph homomorphism is more commonly known as a \emph{subgraph isomorphism}, a
\emph{locally injective homomorphism} is one which is injective when restricted to any individual
vertex and its neighbourhood \cite{DBLP:journals/csr/FialaK08}, and a homomorphism which is
surjective on edges is known as an \emph{epimorphism} \cite{DBLP:journals/dam/GayFMSS14}. Finding
any of these kinds of homomorphism between two given graphs is \NP-complete.  The subgraph
isomorphism problem arises in many practical applications, including ??, and as such there has been
a lot of research into finding algorithms which perform well empirically. Most approaches are based
either upon very fast but simple backtracking algorithms,
\cite{DBLP:journals/pami/CordellaFSV04,DBLP:journals/bmcbi/BonniciGPSF13,DBLP:conf/gbrpr/CarlettiFSV17}
which often but not always perform well on very easy instances, or upon constraint programming
algorithms
\cite{DBLP:journals/constraints/ZampelliDS10,DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/cpaior/ArchibaldDHMP019},
which have higher startup costs but that perform vastly better on harder instances and much more
consistently on easy instances \cite{DBLP:journals/jair/McCreeshPST18,DBLP:conf/gbrpr/Solnon19}. The
current state of the art is the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020}, which is a
dedicated constraint programming solver for subgraph-finding problems. Much of its performance comes
from inference strategies based upon degrees and neighbourhood degree sequences
\cite{DBLP:journals/constraints/ZampelliDS10}, counting paths between vertices
\cite{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15}, and cardinality reasoning
\cite{DBLP:journals/ai/Solnon10}, which can be used to eliminate many infeasible candidate
assignments without search.

One might hope that this inference would also be helpful for other graph homomorphism problem
variants. Indeed, this paper proves that many, but not all, of these strategies are also valid in
the locally injective case, and that simple distance filtering is valid for all homomorphisms.
However, we also show that \emph{none} of the other strategies are valid for finding homomorphisms
where there is no injectivity requirement. Finally, we introduce a new filtering technique that is
based upon finding a maximum clique in the neighbourhood of each domain vertex, and prove that this
technique is valid even in the general case. Although this new filtering technique involves solving
many additional \NP-complete problems as a preprocessing step, we demonstrate that it is effective
in practice, particularly for the non-injective problem where filtering allows a constraint
programming algorithm to solve instances over eight hundred times faster in aggregate, over a
collection of over fourteen thousand standard benchmark instances. This shows, for the first time,
that modern constraint programming techniques can be practical for less constrained graph
homomorphism finding problems; previous algorithmic approaches have focused instead upon worst-case
computational bounds
\cite{DBLP:journals/csr/FialaK08,DBLP:journals/ipl/Rzazewski14,DBLP:journals/tcs/ChaplickFHPT15},
whose practical utility has yet to be demonstrated. The empirical effectiveness of our results is
especially important because many current applications use subgraph isomorphism solvers only because
they perform well using off the shelf solvers, rather than because they exactly match ideal domain
requirements ??cite.

@@

We begin by introducing notation and terminology, and by giving the theoretical results which
support our implementation.

@@

\begin{proposition}[neighbourhood degree sequences are preserved]\label{proposition:degreends}
    Suppose  $i$ is a locally injective homomorphism from a $G$ to  $H$. Let
    $v$ be any vertex. Then:
    \begin{enumerate}
        \item $i$ cannot map $v$ to a vertex of lower degree, $\degree_G(v) \le \degree_H(i(v))$.
        \item Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own.
    \end{enumerate}
\end{proposition}
Since subgraph isomorphisms are in particular locally injective, these invariants also hold for
subgraph isomorphisms \cite{DBLP:journals/constraints/ZampelliDS10}; indeed, every recent constraint
programming approach for subgraph isomorphism finding make use of degrees and neighbourhood degree
sequences. The simplest way for a constraint programming algorithm to use these results is as unary
constraints, which are propagated when domains are initialised, before search begins. For each
domain representing a domain vertex, any value representing a codomain vertex whose degree or
neighbourhood degree sequence is too low may immediately be rejected.

@@

However, these invariants do \emph{not} hold for homomorphisms in general. As a simple
counter-example, a homomorphism may map a star graph onto a single edge, as in
\cref{figure:counterexamples} (top left).

@@

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two (see \cref{figure:counterexamples}, top right). Local injectivity is trickier.
The second property does not hold in full generality, because, for example, a path of length three
may be mapped onto a triangle whilst respecting local injectivity (\cref{figure:counterexamples},
bottom left).  However, we observe that a weaker result holds:

@@

\begin{proof}Let $\{ x_1, \ldots, x_n \}$ be the intermediate vertices on the paths of length two
between $v$ and $w$. Observe that each $x_j$ is in the neighbourhood of $v$, and so must be mapped
    to different vertices. Thus each sequence $(i(v), i(x_j), i(w))$ gives a distinct simple path of
length two between $i(v)$ and $i(w)$ as required.\end{proof}

@@

\begin{proposition}[Cliques are preserved]\label{proposition:clique}
    Let $i$ be a homomorphism from $G$ to $H$ where $H$ does
    not contain any loops. Let $S\subset V$ be such that $G[S]$ is a $k$-vertex clique. Then
    $H[i(S)]$ is also a $k$-vertex clique.
\end{proposition}

@@

We will discuss how this is useful in practice in the following section.

@@

Having determined which commonly-used subgraph isomorphism invariants do and do not hold for other
forms of homomorphism, and having discovered a new clique-based invariant, we will now look at how
these invariants may be used in practice. Our starting point is the Glasgow Subgraph Solver
\cite{DBLP:conf/gg/McCreeshP020}, due to it being the current single strongest solver for the
subgraph isomorphism problem. Adapting the solver to handle locally injective and general
homomorphism problems required the following straightforward changes.

@@

\paragraph{Path-based filtering.} For subgraph isomorphism, the Glasgow Subgraph Solver
searches for a mapping which is simultaneously a subgraph isomorphisms between $(G, H)$ and each
$(G^{n,2}, G^{n,2})$ for each $n$ from $1$ to $4$, and optionally also between $(G^3, H^3)$.

\paragraph{Degree-based filtering.} This was disabled for homomorphisms and left enabled for locally
injective homomorphisms, as per \cref{proposition:degreends}. For subgraph isomorphism, the Glasgow
Subgraph Solver uses degree and neighbourhood degree sequences not just on the original graphs, but
also on each $(G^{n,\ell}, H^{n,\ell})$ graph pair. This is \emph{not} possible for locally
injective homomorphisms, due to the counter-example following \ref{corollary:lishapes}.

@@

We then move on to the codomain vertices, using a similar caching routine. Rather than calculating a
clique size for every single codomain vertex, we only calculate a value for codomain vertices which are
present in at least one variable's domain, after other unary constraints have been applied.
(Depending upon the problem variant, implied constraints like degree and neighbourhood degree
sequence can eliminate some codomain vertices from the domains of all domain variables.)
Additionally, we do not require the maximum clique solver to run to completion and guarantee that it
has found a maximum clique. Instead, we allow it to stop as soon as it has found a clique with as
many vertices as the largest domain clique. This is useful in practice because it may be very hard
to decide whether a particular codomain vertex has neighbourhood clique size of, say, 15 or 16, but if
the largest domain vertex has a neighbourhood clique size of only 5 then this is irrelevant.

?? Why this works. Clique finding is much easier, colour bounds, etc. Small graphs. ?? See in
section blah that it's really fast.

@@

Given the complexity of implementations of modern solvers, a critical question is how we can be sure
that they are producing correct answers. The Glasgow Subgraph Solver's subgraph isomorphism and
clique algorithms are both \emph{certifying} \cite{DBLP:journals/csr/McConnellMNS11}: that is, they
can output a mathematical proof that they have reached a correct answer by sound reasoning, and this
proof log can be verified by the VeriPB proof checking tool
\cite{DBLP:conf/cp/GochtMMNPT20,DBLP:conf/ijcai/GochtMN20}. Certification involves two stages:
firstly, the solver translates its inputs into a pseudo-Boolean model (this part is not verified),
and secondly (for unsatisfiable instances), the solver provides a proof using cutting planes rules
and reverse unit propagation steps that the pseudo-Boolean model is unsatisfiable.

Implementing proof-logging for the changes not involving clique-finding was routing---although proof
logging did catch an implementation bug where we were incorrectly applying degree-based filtering on
the $G^{2,n}$ graphs for locally injective homomorphisms, in violation of the counter-example to
\cref{corollary:lishapes}. For clique-based filtering, more work was required. We briefly
outline the steps taken to implement proof logging for the fact that a vertex $p$ in the domain
graph which is present in a $k$-vertex clique cannot be mapped to a vertex $t$ in the codomain graph
which is not present in a $k$-vertex clique.

?? Many technicalities.

We have an $x_{p{,}t}$ saying that $p$ gets mapped to $t$. For a regular clique problem, we would have
binary constraints $\overline{x}_i + \overline{x}_j \ge 1$ for each non-adjacent vertex $i$ and
$j$. We wish to derive something similar for the non-adjacent vertices in the neighbourhood of $t$.
However, our constraints are conditional upon $p$ being mapped to $t$, and so instead we start our
proof step by deriving three sets of conditional\footnote{For the subgraph isomorphism case, these
constraints need not be conditional.} pseudo-adjacency constraints:
\begin{itemize}
    \item $\overline{x}_{p{,}t} + \overline{x}_{q{,}u} + \overline{x}_{r{,}v} \ge 1$
    \item $\overline{x}_{p{,}t} + \overline{x}_{q{,}u} + \overline{x}_{r{,}u} \ge 1$
    \item $\overline{x}_{q{,}u} + \overline{x}_{q{,}v} \ge 1$
\end{itemize}
Each of these constraints follows by reverse unit propagation from the original model, and so
requires no further justification in a proof log. We also need an objective constraint. From the
definition of the homomorphism problem, we have existing adjacency constraints $\overline{x}_{p{,}t}
+ \sum_{u \in \neighbourhood(t)} x_{q{,}u} \ge 1$, and we may sum these constraints for each of the
$k - 1$ vertices in the $k$-clique that includes $p$ to express that a suitable $k$-clique must
exist in the codomain graph.

We have now established something that is sufficiently similar to the maximum clique model used by
?? that the . Colour classes (used for bounds) are in correspondence with colour classes in this
adapted problem instance, which is similar to microstructure. Produce a proof that there is no
target clique. The condition appears in the trail, and rather than deriving a contradiction, gets
$\overline{x}_{p,t} \ge 1$.

This situation is not entirely ideal. The proofs produced this way are inherently dependent upon a
\emph{particular} choice of $p$ and $t$ vertices. When not proof logging, we need to solve a single
maximum clique problem for each vertex in the domain, and a restricted clique problem for each
vertex in the codomain. With proof logging, we must instead solve restricted clique problems for
each $(p, t)$ assignment that is to be filtered, which is potentially a linear factor blowup. This
feels morally excessive, since for any given $t$ (and for any clique in the codomain graph which is
reused) the proofs produced are effectively the same up to a substitution of variable names. We
would therefore be interested in seeing a proof system extension which would allow us to avoid this
duplication.

@@

    \caption{On the top left, the cumulative number of instances solved over time for the three
    problem variants, with and without clique filtering, and with and without distance filtering for
    the homomorphism problem; for comparison, results for subgraph isomoprhism using the PathLAD,
    RI, and VF2 solvers are also shown. The remaining plots re-display this data, as follows. The
    three plots on the bottom row zoom in on the cumulative number of instances solved, for the
    homomorphism problem on the left, the locally injective homomorphism problem in the centre, and
    the subgraph isomorphism problem on the right. The top centre plot shows the additional number
    of instances solved at any given time when using the new forms of filtering for all problem
    variants, and the top right plot shows the aggregate speedups from each form of filtering.}
\end{figure*}
\begin{figure*}[p]
    \,\hfill\begin{tabular}{c@{\hspace{4mm}}c@{\hspace{4mm}}c@{\hspace{4mm}}c@{\hspace{4mm}}}
        \includegraphics{gen-graph-scatter-hom-c.pdf}
        &
        \includegraphics{gen-graph-scatter-hom-cd.pdf}
        &
        \includegraphics{gen-graph-scatter-local.pdf}
        &
        \includegraphics{gen-graph-scatter-si.pdf}
    \end{tabular}\hfill\,

    \caption{Looking at the effects of additional filtering on an instance by instance basis, for
    homomorphism with just clique filtering and with both clique and distance filtering, and for the
    other two variants with clique filtering. Each point represents one instance, the vertical axis
    is the runtime with filtering in ms, and the horizontal axis is the runtime without filtering in
    ms (and so points below the diagonal are speedups). Points on the outer axes are timeouts. The
    different point styles show the different families of instance from the benchmark set, and
    illustrate that in each the filtering is broadly useful rather than being specific to a single
    kind of application.}
\end{figure*}

We now evaluate this approach empirically on

?? How good is it

?? Time on cliques

?? Whether cliques are hard

?? Loops

\section{Conclusion and Future Work}

?? Neighbourhood colourability? Cores?

?? Epimorphism

\section*{Acknowledgements}

This work was supported by the Engineering and Physical Sciences Research Council [grant number
EP/P026842/1].

>>>>

% \usepackage[textsize=tiny,textwidth=2cm,shadow]{todonotes}
% \newcommand{\skcom}[1]{\todo[color=orange!50!white]{{S}: #1}}
% \newcommand{\rcom}[1]{\todo[color=gray!50!white]{{R}: #1}}%reviewers said
% \usepackage{showframe}

@@

\author{Paper 4726}

% \author{
% Ciaran McCreesh$^1$
% \and
% Sonja Kraiczy$^2$
% \affiliations
% $^1$University of Glasgow\\
% $^2$University of Oxford\\
% \emails
% ciaran.mccreesh@glasgow.ac.uk
% }

@@

The \emph{subgraph isomorphism problem}, as defined by \citet{DBLP:books/fm/GareyJ79}, is to
determine whether an injective mapping exists from one given graph to another, such that adjacent
pairs of vertices are mapped to adjacent pairs of vertices, and non-adjacent vertices are mapped to
non-adjacent vertices.  However, in application-oriented papers, particularly from bioinformatics
and chemistry, the same name is often implicitly used to mean the \emph{non-induced} version of the
problem (which does not require that non-adjacency be preserved) since this more accurately models
the real-world problem being solved \cite{Willett1999,EhrlichR11}. Despite being theoretically hard
problems, these applications, along with
others in areas including compilers \cite{DBLP:conf/cp/BlindellLCS15}
and pattern recognition \cite{DBLP:journals/ijprai/FoggiaPV14},
have given rise to a large amount of research into designing practical algorithms for solving these
problems.  Most approaches are based either upon very fast but simple backtracking algorithms
\cite{DBLP:journals/pami/CordellaFSV04,DBLP:journals/bmcbi/BonniciGPSF13,DBLP:conf/gbrpr/CarlettiFSV17}
which often but not always perform well on very easy instances, or upon constraint programming
algorithms
\cite{DBLP:journals/constraints/ZampelliDS10,DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/cpaior/ArchibaldDHMP019},
which have higher startup costs but that perform vastly better on harder instances and much more
consistently on easy instances \cite{DBLP:journals/jair/McCreeshPST18,DBLP:conf/gbrpr/Solnon19}.

The current state of the art is the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020}, which
is a dedicated constraint programming solver for subgraph-finding problems. Much of its performance
comes from inference strategies based upon degrees and neighbourhood degree sequences
\cite{DBLP:journals/constraints/ZampelliDS10}, counting paths between vertices
\cite{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15}, and cardinality reasoning
\cite{DBLP:journals/ai/Solnon10}, which can be used to eliminate many infeasible candidate
assignments without search. Like most other solvers, it can handle both the non-induced version of
the problem, as well as the traditional induced variant, and users can select whichever variant
better fits their application. It also supports directed edges, and can use an external constraint
solver to handle other extensions to the basic notion of a graph, such as temporal networks and
multi-graphs.

As well as questions over exactly how edges are represented and mapped, some applications would
prefer not to specify injectivity. A \emph{graph homomorphism} is a function between two graphs that
maps adjacent vertices to adjacent vertices, with no injectivity requirement, whilst a \emph{locally
injective homomorphism} is one which is injective when restricted to any individual vertex and its
neighbourhood. Both are better models for some applications
\cite{DBLP:journals/dam/FialaKK01,DBLP:conf/semweb/Baget05,DBLP:conf/iccs/CorbyF07,DBLP:journals/csr/FialaK08,DBLP:journals/pvldb/FanLMWW10,DBLP:journals/tcs/ChaplickFHPT15}.
(There is also an edge-surjective version of the problem known as an \emph{epimorphism}
\cite{DBLP:journals/dam/GayFMSS14}.)

These injectivity relaxations have received much less attention
from a practical solving perspective.  One might hope that the inference techniques developed for
subgraph isomorphism would also be helpful for other graph homomorphism problem variants. Indeed,
this paper proves that many, but not all, of these strategies are also valid in the locally
injective case, and that simple distance filtering is valid for all homomorphisms.  However, we also
show that \emph{none} of the other strategies are valid for finding homomorphisms where there is no
injectivity requirement.  Finally, we introduce a new filtering technique that is based upon finding
a maximum clique in the neighbourhood of each domain vertex, which is valid even in the general
case. Although this new filtering technique involves solving many additional \NP-complete problems
as a preprocessing step, we demonstrate that it is effective in practice, particularly for the
non-injective problem where filtering allows a constraint programming algorithm to solve instances
over eight hundred times faster in aggregate, over a collection of over fourteen thousand standard
benchmark instances. This shows, for the first time, that modern constraint programming techniques
can be practical for less constrained graph homomorphism finding problems; previous algorithmic
approaches have focused instead upon worst-case computational bounds
(e.g.\
\citet{DBLP:journals/mst/FominHK07,DBLP:journals/csr/FialaK08,DBLP:journals/ipl/Rzazewski14,DBLP:journals/tcs/ChaplickFHPT15}),
whose practical utility has yet to be demonstrated. The empirical effectiveness of our results is
especially important because many current applications use subgraph isomorphism solvers only because
they perform well using off the shelf solvers, rather than because they exactly match ideal domain
requirements.

@@

We begin by introducing notation and terminology, and by giving the theory which
support our implementation.

@@

\begin{proposition}[neighbourhood degree sequences are preserved]\label{proposition:degreends}
    Suppose  $i$ is a locally injective homomorphism from a $G$ to  $H$. Let
    $v$ be any vertex. Then $i$ cannot map $v$ to a vertex of lower degree, $\degree_G(v) \le \degree_H(i(v))$.
    Furthermore, $v$ cannot be mapped to any vertex whose neighbourhood degree sequence is
            not pointwise greater than or equal to its own.
\end{proposition}
Since subgraph isomorphisms are in particular locally injective, these invariants also hold for
subgraph isomorphisms \cite{DBLP:journals/constraints/ZampelliDS10}; indeed, every recent constraint
programming approach for subgraph isomorphism finding make use of degrees and neighbourhood degree
sequences. The simplest way for a constraint programming algorithm to use these results is as unary
constraints, which are propagated when domains are initialised, before search begins. For each
domain representing a domain vertex, any value representing a codomain vertex whose degree or
neighbourhood degree sequence is too low may immediately be rejected.

@@

However, these invariants do \emph{not} hold for homomorphisms in general. As a
counter-example, a homomorphism may map a star graph onto a single edge, as in
\cref{figure:counterexamples} (top left).

@@

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two (see \cref{figure:counterexamples}, top right). While standard injectivity
prevents any kind of merging, local injectivity prevents two vertices from being merged if they are
both in the neighbourhood of the same vertex.  So the second property does not hold in full
generality, because, for example, a path of length three may be mapped onto a triangle whilst
respecting local injectivity (\cref{figure:counterexamples}, bottom left).  However, we observe that
a weaker result holds:

@@

\begin{proof}Let $\{ x_1, \ldots, x_n \}$ be the intermediate vertices on the paths of length two
between $v$ and $w$. Observe that each $x_j$ is in the neighbourhood of $v$, and so must be mapped
    to different vertices. Thus each sequence $(i(v), i(x_j), i(w))$ gives a distinct simple path of
length two between $i(v)$ and $i(w)$.\end{proof}

@@

\begin{proposition}[Cliques are preserved]\label{proposition:clique}
    Let $i$ be a homomorphism from $G$ to $H$ where $H$ does
    not contain any loops. Let $S\subseteq V$ be such that $G[S]$ is a $k$-vertex clique. Then
    $H[i(S)]$ is also a $k$-vertex clique.
\end{proposition}

@@

Next we will discuss how this is useful in practice.

@@

Having determined which commonly-used subgraph isomorphism invariants do and do not hold for other
forms of homomorphism, and having discovered a new clique-based invariant, we will now look at how
these invariants may be used in practice. Our starting point is the Glasgow Subgraph Solver
\cite{DBLP:conf/gg/McCreeshP020}, due to it being the current strongest solver for
subgraph isomorphism. Adapting the solver to handle locally injective and general homomorphism
problems required the following straightforward changes.

@@

\paragraph{Path-based filtering.} For subgraph isomorphism, the Glasgow Subgraph Solver
searches for a mapping which is simultaneously a subgraph isomorphisms between $(G, H)$ and each
$(G^{n,2}, G^{n,2})$ for each $n$ from $1$ to $4$, and optionally also between $(G^3, H^3)$. For
homomorphisms, we use $(G^2, H^2)$, whilst for locally injective homomorphisms we use $(G^{n,2},
G^{n,2})$ for each $n$ from $1$ to $4$.

\paragraph{Degree-based filtering.} This was disabled for homomorphisms and left enabled for locally
injective homomorphisms, as per \cref{proposition:degreends}. For subgraph isomorphism, the Glasgow
Subgraph Solver uses degree and neighbourhood degree sequences not just on the original graphs, but
also on each $(G^{n,\ell}, H^{n,\ell})$ graph pair. This is \emph{not} possible for locally
injective homomorphisms, due to the counter-example following \cref{corollary:lishapes}.

@@

We then move on to the codomain vertices, using a similar caching routine. Rather than calculating a
clique size for every single codomain vertex, we only calculate a value for codomain vertices which are
present in at least one variable's domain, after other unary constraints have been applied.
Additionally, we do not require the maximum clique solver to run to completion and guarantee that it
has found a maximum clique. Instead, we allow it to stop as soon as it has found a clique with as
many vertices as the largest domain clique. This is useful in practice because it may be very hard
to decide whether a particular codomain vertex has neighbourhood clique size of, say, 15 or 16, but if
the largest domain vertex has a neighbourhood clique size of only 5 then this is irrelevant.

@@

Given the complexity of implementations of modern solvers, a critical question is how we can be sure
that they are producing correct answers. The Glasgow Subgraph Solver's subgraph isomorphism and
clique algorithms are both \emph{certifying} \cite{DBLP:journals/csr/McConnellMNS11}: that is, they
can output a mathematical proof that they have reached a correct answer by sound reasoning, and this
proof log can be verified by the VeriPB proof checking tool
\cite{DBLP:conf/cp/GochtMMNPT20,DBLP:conf/ijcai/GochtMN20}.

Implementing proof-logging for the changes not involving clique-finding was routine---although proof
logging did catch an implementation bug where we were incorrectly applying degree-based filtering on
the $G^{2,n}$ graphs for locally injective homomorphisms, in violation of the counter-example to
\cref{corollary:lishapes}. For clique-based filtering, the process involved substantially more
technical work. Briefly, for a given domain vertex $v$ that cannot be mapped to a codomain vertex
$w$, it is possible to instruct the proof verifier to derive a set of conditional clique constraints
saying that ``either $v$ cannot be mapped to $w$, or the following clique-like problem must be
satisfiable'' by following steps similar to the maximum common subgraph to clique reduction
presented by \citet{DBLP:conf/cp/GochtMMNPT20}, and then to reuse clique
proof-logging techniques to show that this clique-like problem is unsatisfiable.

This situation is not entirely ideal. The proofs produced this way are inherently dependent upon a
\emph{particular} choice of domain and codomain vertices. When not proof logging, we only need to
solve a single maximum clique problem for each vertex in the domain, and a restricted clique problem
for each vertex in the codomain. With proof logging, we must instead solve restricted clique
problems for each assignment that is to be filtered, which is potentially a linear factor blowup.
This seems wasteful, since for any given codomain vertex these proofs are effectively the same up to
substitution of variable names. It would therefore seem desirable to extend the proof system with a
substitution rule which would allow such proofs to be recycled to avoid duplication.

@@

    \caption{On the top left, the cumulative number of instances solved over time for the three
    problem variants, with and without clique filtering, and with and without distance filtering for
    the homomorphism problem; for comparison, results for subgraph isomoprhism using the PathLAD,
    RI, and VF2 solvers are also shown. The remaining plots re-display this data, as follows. The
    three plots on the bottom row zoom in on the cumulative number of instances solved, for the
    homomorphism problem on the left, the locally injective homomorphism problem in the centre, and
    the subgraph isomorphism problem on the right. The top centre plot shows the additional number
    of instances solved at any given time when using the new forms of filtering for all problem
    variants, and the top right plot shows the aggregate speedups from each form of
    filtering.}\label{figure:plots}
\end{figure*}
\begin{figure*}[p]
    \,\hfill\begin{tabular}{c@{\hspace{4mm}}c@{\hspace{4mm}}c@{\hspace{4mm}}c@{\hspace{4mm}}}
        \includegraphics{gen-graph-scatter-hom-c.pdf}
        &
        \includegraphics{gen-graph-scatter-hom-cd.pdf}
        &
        \includegraphics{gen-graph-scatter-local.pdf}
        &
        \includegraphics{gen-graph-scatter-si.pdf}
    \end{tabular}\hfill\,

    \caption{Looking at the effects of additional filtering on an instance by instance basis, for
    homomorphism with just clique filtering and with both clique and distance filtering, and for the
    other two variants with clique filtering. Each point represents one instance, the vertical axis
    is the runtime with filtering in ms, and the horizontal axis is the runtime without filtering in
    ms (and so points below the diagonal are speedups). Points on the outer axes are timeouts. The
    different point styles show the different families of instance from the benchmark set, and
    illustrate that in each the filtering is broadly useful rather than being specific to a single
    kind of application.}\label{figure:scatters}
\end{figure*}

We now evaluate this approach empirically. Our experiments are performed on a cluster of machines
with dual Intel Xeon E5-2697A v4 CPUs with 512GBytes of RAM, running Ubuntu 18.04, and using a
timeout of one hour. Because injectivity relaxations have received less attention until now, for all
three problem variants we will use a suite of 14,621 benchmark instances designed for the subgraph
isomorphism problem \cite{DBLP:conf/lion/KotthoffMS16}. Note that 8,904 of these instances contain
at least one loop in the codomain graph, and so are trivial for the homomorphism problem---we still
include these to avoid using different y-axes on different plots.

In the top left plot of \cref{figure:plots} we show the cumulative number of instances solved over
time, for each of the three problem variants, using the Glasgow Subgraph Solver with and without our
clique-filtering added. For the homomorphism problem, we also compare with distance filtering
enabled or disabled. For subgraph isomorphism, we also compare against the PathLAD
\cite{DBLP:conf/lion/KotthoffMS16}, VF2 \cite{DBLP:journals/pami/CordellaFSV04} and RI
\cite{DBLP:journals/bmcbi/BonniciGPSF13} solvers. Because every Glasgow Subgraph Solver result is so
much stronger than the other three solvers, it is hard to get any further information from this
plot. We therefore re-plot zoomed in results on the second row of \cref{figure:plots}, focusing each
problem variant in turn. The results show that for the homomorphism problem, using either clique or
distance filtering gives a clear improvement to performance, and using both together is better
still. For locally injective homomorphisms and subgraph isomorphism, clique filtering also gives an
improvement.

To quantify these gains, the top centre plot of \cref{figure:plots} shows the
additional number of instances solved at any given time (that is, the vertical distance between
two cumulative plots) when enabling additional filtering. We see that for subgraph isomorphism, at
the one hour timeout we solve 22 additional instances using clique filtering; for locally injective
homomorphisms, 21 additional instances; and for homomorphisms, 114, 146 or 193 additional instances
using just distance filtering, just clique filtering, or both. Alternatively, we can measure the
horizontal distance between plots for a given choice of timeout $t$ as follows. Let $t'$
be the last time no greater than $t$ where the slower algorithm solved an instance, and let $u$ be
the lowest amount of time needed for the faster algorithm to solve the same number of instances.
Then the aggregate speedup, which we plot in the top right of \cref{figure:plots}, is the ratio of
$t'$ to $u$ or $u$ to $t'$ as appropriate. At the timeout, we see aggregate speedups of 2.2 for
locally injective homomorphisms, 3.2 for subgraph isomorphism, and 47.5, 131.7, and 803.5 for
homomorphisms.

The crossover point, where clique filtering begins to pay off, is at around the 100ms mark for all
homomorphism variants, and at around one second for subgraph isomorphism. This does not mean that
this much time was spent in clique-finding, but rather that this is how long it takes for an
algorithm with increased filtering to catch up from a slower start. In fact, for domain graphs, no
maximum clique call took even one millisecond to complete (and none required more than two hundred
recursive calls from the clique solver), and more time was spent setting up the necessary data
structures to represent the graphs than on the solving itself. For codomain graphs, 20,838,008 of
the clique problems we solved took under a millisecond, and the remaining 279 calls took between one
and four milliseconds, with no instance requiring more than a thousand recursive calls to solve.
Although it may seem counter-intuitive to try to speed up the solving of a hard problem by first
solving many other hard problems, these values demonstrate that clique-solving on these smaller
graphs is \emph{much} easier in practice than solving the larger problem, and so is worthwhile.

Finally, \cref{figure:scatters}, plots instance-by-instance comparisons of the runtimes. The
point styles in these plots show different families of benchmark instance, and demonstrate
that the benefits of filtering are not restricted to one application.

\section{Conclusion}

These results show that constraint programming approaches are viable for a wider range of subgraph
mapping finding problems than had previously been considered, which will increase the versatility of
subgraph solvers for application users going forwards. We also saw that using clique-finding as a
filtering step could give moderate to very large speedups for different problem variants---although
only through careful engineering choices, to mitigate the potential cost of solving \NP-hard
problems. We would be interested to see if filtering based upon other hard problems could give
similar benefits---for example, homomorphisms also interact in potentially helpful ways with various
colouring properties.

% \section*{Acknowledgements}
% 
% This work was supported by the Engineering and Physical Sciences Research Council [grant number
% EP/P026842/1].
